Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2015-02-17 parts/ccl.tex
finish corrections

In this paper, we focused on the Asynchronous Automata Networks (AANs),
which is a restriction of classical Automata Networks
and is equivalent to the Process Hitting with classes of priorities~\cite{FPMR13-CS2Bio}.
This formalism proves useful to model accurate Boolean gates, which was not possible
with the standard form of the Process Hitting,
and avoid an unwanted over-approximation of the dynamics.
Furthermore, we proposed an extension of this formalism with classes of priorities,
which prove convenient to abstract time parameters into these kinds of models,
or more simply to add preemption relations between actions.

Then, we developed a method to perform a reachability analysis
of a sequence of objectives in an AAN,
based on an under-approximation of the true reachability solutions.
The checked objectives can take the form of local states, global states
or partial states, making the analysis more convenient.
This method can be considered efficient,
because it is exponential in the number of possible solutions,
and each step is only polynomial in the size of the model.
Furthermore, AANs with classes of priorities can also be studied in this way,
at the cost of a translation that we gave in this paper.

Finally, we also showed that AANs are also equivalent to Asynchronous Discrete Networks,
that is, the multivalued version of the classical Asynchronous Boolean Networks
with discrete expression levels and evolution functions.
This framework is also equivalent to Thomas' models with Snoussi parameters
instead of evolution functions.
This especially allows to efficiently compute reachability results
on large biological models,
provided that they are equivalent to Asynchronous Discrete Networks,
which ensures that a translation to AANs is possible.
For example,
such a translation for generalized Interaction Graphs,
that is, Discrete Networks without evolution functions or parameters,
was proposed in~\cite{PMR10-TCSB}.

Further work can be derived from what have been presented in this paper.
The over-approximation on Process Hitting models without priorities proposed in~\cite{PMR12-MSCS}
is still accurate in the framework with priorities (by “merging” all actions),
but may be refined given the restrictions proposed in this paper,
and a specific search of key processes or cut sets may be derived.
%but turns out to be too wide even in some obvious cases that are consequently not conclusive.
%This approximation may be refined in order to better fit the introduction of priorities, and mak the overall approximation approach 
%and mode precisely the class of models studied in this paper.
Furthermore, we are investigating alternative under-approximations that can be
applied directly to the whole class of Process Hitting models with priorities,
and not only to a sub-class with particular restrictions;
such improvement may allow to increase the conclusiveness of the static analysis
while allowing to analyse any model without the need of a translation.
Finally, in order to take into account quantitative data in transition delays, the overall approximation method could be extended to handle evolutions that are chronometric instead of only chronologic.

>>>>

In this paper, we focused on Asynchronous Automata Networks (AANs),
which are a restriction of classical Automata Networks
and are equivalent to Process Hitting models with classes of priorities~\cite{FPMR13-CS2Bio}.
This formalism proves useful to model accurate Boolean gates, which was not possible
with the standard form of the Process Hitting,
and avoid an unwanted over-approximation of the dynamics.
Furthermore, we proposed an extension of this formalism with classes of priorities,
which prove convenient to abstract time parameters into these kinds of models
or more simply to add preemption relations between actions,
and showed that any AAN with priorities can be translated into
an equivalent AAN without priorities.

Then, we developed a method to perform a reachability analysis
of a local state in an AAN,
based on an under-approximation of the true reachability solutions.
We also extended this analysis to global and
partial states, and to the successive reachability of several local states.
The method can be considered efficient,
because it is polynomial in the size of the model.
A more conclusive analysis also exists,
but at the price of being exponential in the number of local solutions.
Finally, AANs with classes of priorities can also be studied in this way,
at the cost of an exponential translation that we gave in this paper.
We applied it to a large number of experiments on two large-scale
biological models, and obtained in the worst case a ratio of
%65\% of positively responding cases, growing to
90\% of conclusive cases
with the joint use of a previously proposed over-approximation,
although limiting the computation time to 3 seconds for each test.

AANs are also equivalent to Logical Networks, that is,
either multivalued or Boolean networks
with evolution functions or focal parameters,
such as Thomas' models with Snoussi parameters.
This especially allows to efficiently compute reachability results
on large biological models,
provided that they are equivalent to Logical Networks,
which ensures that a translation to AANs is possible.
For example,
such a translation for generalized Interaction Graphs,
that is, Discrete Networks without evolution functions or parameters,
was proposed in~\cite{PMR10-TCSB}.

Further work can be derived from what have been presented in this paper.
The over-approximation on Process Hitting models without priorities proposed in~\cite{PMR12-MSCS}
and that was used in this work
is still accurate on AANs (by over-approximating dynamics)
but may be refined given the particular for of AANs proposed in this paper.
A specific search of key processes or cut sets \cite{PAK13-CAV} may especially be derived.
%but turns out to be too wide even in some obvious cases that are consequently not conclusive.
%This approximation may be refined in order to better fit the introduction of priorities, and mak the overall approximation approach 
%and mode precisely the class of models studied in this paper.
Furthermore, we are investigating alternative under-approximations that can be
applied directly to the whole class of AANs or Process Hitting models with priorities,
and not only to a sub-class with only one class of priority or particular restrictions;
such improvement may allow to increase the conclusiveness of the static analysis
while allowing to analyse any model without the need of a translation.
Finally, in order to take into account quantitative data in transition delays, the overall approximation method could be extended to handle evolutions that are chronometric instead of only chronologic.

### 2015-02-17 parts/example.tex
finish corrections

The TCR signalling pathway consists of 94 components.
We checked the reachability for the independent activation of
the 4 outputs of the signalling cascade (SRE, AP1, NFkB, NFAT)
for all possible combinations of the 3 inputs (CD4, CD28, TCRlig).
All result in conclusive decisions,
and our under-approximation has been satisfied in 12 cases (over
32) proving the satisfiability of the concerned reachability property in the encoded Boolean network
(and non-satisfiability in the other cases, using the previously existing over-approximation).

By analysing the detail of the results, one can find out that one of the outputs, NFkB, can never be activated.
In other words, FkB$_1$ is never reachable, whatever the initial state of the inputs,
while the other outputs (SRE, AP1, NFAT) can be independently activated
for some configurations of inputs.
We thus want to check if these three outputs can be activated together,
that is, if the activation of one of the outputs does not prevent the activation another one.
For this, as proposed in \pref{ssec:simult-ua},
we add a new automaton $\reach$ into the model
with two local states ($\PHl_\reach = \{ \reach_0, \reach_1 \}$)
and the action:
\[ \PHhits{\text{SRE}_1, \text{AP1}_1, \text{NFAT}_1}{\reach_0}{\reach_1} \enspace. \]
Finally, checking the reachability of $\reach_1$ in all configurations of the inputs
is always conclusive,
and the existence of 4 positive answers
(amongst all 8 possible initial configurations of the inputs) allows to conclude that
it is possible to reach a state where SRE, AP1 and NFAT are simultaneously active.

@@

The EGFR signalling pathway gathers 104 components, amongst which one can distinguish
21 inputs and 12 outputs.
Analysing the consequences of the variations of all inputs
would require too many tests;
therefore, we selected 13 of them
(erbb1, erbb2, erbb3, erbb4, bir, btc, egf, epr, nrg1a, nrg1b, nrg2b, nrg4, tgfa)
and observed the impact of their variations on all outputs
(elk1, creb, ap1, hsp27, actin\_reorg, cmyc, pro\_apoptotic, p70s6\_2, pkc, stat1, stat3, stat5).
We note that some of the experiments trigger the exponential search
of sub-solutions described above, and are cut after 3 seconds of computation
(thus leading to \emph{Inconclusive} cases).
These experiments represent about 10\% of all tests,
amongst the remaining 90\%, while all the others are conclusive.
We note that none of the tests terminated normally (without being cut after 3 seconds)
with an inconclusive answer, which is however theoretically possible.

@@

The results of all tests performed with the implementation of our under-approximation
are summarised in \pref{tab:results} (columns “AAN”).
We held all these experiments on a personal computer,
and computations times were in the order of a few tenths of a second to about one second.
To give a comparison, we did the same experiments with a standard symbolic model-checker, libDDD
\cite{libddd}, known for its good performances, the input model being the Boolean network expressed
as a Petri net.
However, due to the large scale of the model,
this program takes at least several minutes to terminate,
and runs out of memory for the majority of all experiments.
On the other hand,
our method is able to conclude with limited memory and computation time usage
in the majority of the cases,
and is expected to be scalable to models that are even larger,
even by orders of magnitude.

Finally, we note that similar experiments were conducted in~\cite{PMR12-MSCS}.
However, if these experiments allowed to obtain some information
on the TCR and EGFR models to some extent,
they did not provide a formal proof of the reachabilities that have been
checked in the examples above.
Indeed, the semantics of Process Hitting (without classes of priorities)
does not allow to model accurate Boolean gates,
thus leading to some unwanted spurious behaviours
that especially take the form of temporal shifts.
The adding of classes of priorities allows to remove these temporal shifts,
as explained in~\cite{FPMR13-CS2Bio},
with a construction that is equivalent to the AANs presented in \pref{sec:ph}.
Therefore, only the method presented in this paper provides a formal proof that
the observed behaviours are the result of the true dynamics of the systems.
The same experiments have been conducted with the previous version
of the under-approximation and reported alongside in \pref{tab:results}
(in columns “PH”).
We note that indeed, about 10\% of the positive results regarding the EGFR model
could not be proven with our new under-approximation,
and are thus still formally uncertain.

\definecolor{lightgraycell}{gray}{0.8}
\newcommand{\grcl}{\cellcolor{lightgraycell}}

\begin{table}[htp]
  \centering
  \begin{tabular}{|c||c|c||c|c|}
  \hline
    & \multicolumn{2}{c||}{TCR} & \multicolumn{2}{c|}{EGFR} \\
  \hline
  \hline
    Inputs & \multicolumn{2}{c||}{3} & \multicolumn{2}{c|}{13} \\
  \hline
    Outputs & \multicolumn{2}{c||}{4 + $\reach$} & \multicolumn{2}{c|}{12} \\
  \hline
    Total tests & \multicolumn{2}{c||}{40} & \multicolumn{2}{c|}{98'304} \\
  \hline
  \hline
               & PH        & AAN       & PH               & AAN              \\
  \hline
  \hline
    True       & 16 (40\%) & \grcl 16 (40\%) & 74'268 (75,55\%) & \grcl 64'282 (65,39\%) \\
  \hline
    Inconclusive & 0 (0\%) & \grcl 0 (0\%)   & 0 (0\%)          & \grcl 9'986 (10,16\%)  \\
  \hline
    False      & \multicolumn{2}{c||}{24 (60\%)} & \multicolumn{2}{c|}{24'036 (24,45\%)} \\
  \hline
  \hline
    Max time   & 0.043s    & \grcl 0.20s     & 0.37s            & \grcl 0.87s            \\
  \hline
    Total time & <1s       & \grcl <1s       & 45min            & \grcl 9h50min          \\
  \hline
  \end{tabular}
  \caption{\label{tab:results}%
    Results of the tests on large-scale examples.
    The “AAN” column gives the related results on AAN models,
    using the under-approximation presented in this paper,
    while the “PH” column gives the results for PH models
    using cooperative sorts to model actions with multiple hitters,
    and using the under-approximation of~\cite{PMR12-MSCS}.
    The lines labelled “True”, “Inconclusive” and “False” lines give respectively
    the number of positive, cut and negative answers,
    while “Max time” and “Total time” depict respectively
    the maximum time of the individual computations (except those cut after 3 seconds)
    and the overall execution time of all tests (including those cut after 3 seconds).
    The greyed cells highlight the results that are proper to the
    under-approximation presented in this paper.
  }
\end{table}

In conclusion,
while ensuring a low complexity for the analysis of reachability in Boolean and discrete networks,
our under-approximation method turns out to be conclusive in numerous cases when applied to real
large-scale biological models, which were not tractable with exact model-checking.

>>>>

The TCR signalling pathway consists of 94 components.
We checked the reachability for the independent activation of
the 4 outputs of the signalling cascade (SRE, AP1, NFkB, NFAT)
for all possible combinations of the 3 inputs (CD4, CD28, TCRlig).
All result in conclusive decisions,
and our under-approximation has been satisfied in 12 cases (over
32) proving the satisfiability of the concerned reachability properties in the encoded Boolean network
(and non-satisfiability in the other cases, using the previously existing over-approximation).

By analysing the detail of the results, one can find out that one of the outputs, NFkB, can never be activated.
In other words, FkB$_1$ is never reachable, whatever the initial state of the inputs,
while the other outputs (SRE, AP1, NFAT) can be independently activated
for some configurations of inputs.
We thus wanted to check if these three outputs could be activated together,
that is, if the activation of one of the outputs did not prevent the activation another one.
For this, as proposed in \pref{ssec:simult-ua},
we have added a new automaton $\reach$ into the model
with two local states ($\PHl_\reach = \{ \reach_0, \reach_1 \}$)
and the action:
\[ \PHhits{\text{SRE}_1, \text{AP1}_1, \text{NFAT}_1}{\reach_0}{\reach_1} \enspace. \]
Finally, checking the reachability of $\reach_1$ in all configurations of the inputs
has always been conclusive,
and the existence of 4 positive answers
(amongst all 8 possible initial configurations of the inputs) allows to conclude that
it is possible to reach a state where SRE, AP1 and NFAT are simultaneously active.

@@

The EGFR signalling pathway gathers 104 components, amongst which one can distinguish
21 inputs and 12 outputs.
We selected 13 of all inputs
(erbb1, erbb2, erbb3, erbb4, bir, btc, egf, epr, nrg1a, nrg1b, nrg2b, nrg4, tgfa)
and observed the impact of their variations on all outputs
(elk1, creb, ap1, hsp27, actin\_reorg, cmyc, pro\_apoptotic, p70s6\_2, pkc, stat1, stat3, stat5).
We note that some of the experiments trigger the exponential search
of sub-solutions described above, and are cut after 3 seconds of computation
(thus leading to \emph{Inconclusive} cases).
These “interrupted” experiments represent about 10\% of all tests.
%amongst the remaining 90\%, while all the others are conclusive.
We note that amongst all tests that terminated “normally” (without being cut after 3 seconds)
%none of the tests terminated normally (without being cut after 3 seconds)
none of them responded with an inconclusive answer, which is however theoretically possible.

@@

The results of all tests performed with the implementation of our under-approximation
are summarised in \pref{tab:results} (columns “AAN”).
We held all these experiments on a personal computer,
and regarding the experiments that were not cut at the 3 seconds limit,
computations times were in the order of a few tenths of a second to about one second.
To give a comparison, we did the same experiments with a standard symbolic model-checker, LibDDD
\cite{libddd}, known for its good performances, the input model being the Boolean network expressed
as a Petri net.
However, due to the large scale of the model,
this program takes at least several minutes to terminate,
and runs out of memory for the majority of all experiments.
On the other hand,
our method is able to conclude with limited memory and computation time usage
in the majority of the cases,
and is expected to be scalable to models that are even larger,
even by orders of magnitude.

Finally, we note that similar experiments were conducted in~\cite{PMR12-MSCS}.
However, if these experiments allowed to obtain some information
on the TCR and EGFR models to some extent,
they did not provide a formal “True” response regarding the reachabilities that have been
checked in the examples above.
Indeed, the semantics of Process Hitting (without classes of priorities)
does not allow to model accurate Boolean gates,
thus leading to some unwanted spurious behaviours
that especially take the form of temporal shifts.
The adding of classes of priorities allows to remove these temporal shifts,
as explained in~\cite{FPMR13-CS2Bio},
with a construction that is equivalent to the AANs presented in \pref{sec:ph}.
Therefore, only the method presented in this paper provides a formal proof that
the observed behaviours are the result of the true dynamics of the systems.
Thus, additionally to the experiments previously mentioned,
we conducted similar experiments with the previous version
of the under-approximation, that are reported alongside in \pref{tab:results}
(in columns “PH”).
We note that indeed, about 15\% of the positive results regarding the EGFR model
could not be proven with our new under-approximation,
and are thus still formally uncertain.

\definecolor{lightgraycell}{gray}{0.85}
\newcommand{\grcl}{\cellcolor{lightgraycell}}

\begin{table}[htp]
  \centering
  \begin{tabular}{|c||c|c||c|c|}
  \hline
    & \multicolumn{2}{c||}{TCR} & \multicolumn{2}{c|}{EGFR} \\
  \hline
  \hline
    Inputs & \multicolumn{2}{c||}{3} & \multicolumn{2}{c|}{13} \\
  \hline
    Outputs & \multicolumn{2}{c||}{4 + $\reach$} & \multicolumn{2}{c|}{12} \\
  \hline
    Total tests & \multicolumn{2}{c||}{40} & \multicolumn{2}{c|}{98'304} \\
  \hline
  \hline
               & PH        & AAN       & PH               & AAN              \\
  \hline
  \hline
    True       & 16 (40\%) & \grcl 16 (40\%) & 74'268 (75,55\%) & \grcl 64'282 (65,39\%) \\
  \hline
    Inconclusive & 0 (0\%) & \grcl 0 (0\%)   & 0 (0\%)          & \grcl 9'986 (10,16\%)  \\
  \hline
    False      & \multicolumn{2}{c||}{24 (60\%)} & \multicolumn{2}{c|}{24'036 (24,45\%)} \\
  \hline
  \hline
    Max time   & 0.043s    & \grcl 0.20s     & 0.37s            & \grcl 0.87s            \\
  \hline
    Total time & <1s       & \grcl <1s       & 45min            & \grcl 9h50min          \\
  \hline
  \end{tabular}
  \caption{\label{tab:results}%
    Results of the tests on large-scale examples.
    The “AAN” column gives the related results on AAN models,
    using the under-approximation presented in this paper,
    while the “PH” column gives the results for PH models
    using cooperative sorts to model actions with multiple hitters,
    and using the under-approximation of~\cite{PMR12-MSCS}.
    The lines labelled “True”, “Inconclusive” and “False” give respectively
    the number of positive answers,
    experiments cut after 3 seconds and negative answers;
    while “Max time” and “Total time” depict respectively
    the maximum time of the individual computations (except those cut at 3 seconds)
    and the overall execution time of all tests (including those cut at 3 seconds).
    The greyed cells highlight the results that are proper to the
    under-approximation presented in this paper.
  }
\end{table}

In conclusion,
while ensuring a low complexity for the analysis of reachability in Boolean and discrete networks,
our under-approximation method turns out to be conclusive in numerous cases when applied to real
large-scale biological models, which were not tractable in most cases with exact model-checking.
